---
title: 'DATA 606 Data Project Proposal'
author: "Donald Butler"
date: "11/07/2021"
urlcolor: blue
bibliography: references.bib
biblio-style: apalike
link-citations: true
output:
  pdf_document: default
  html_document: default
  word_document: default
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}

library(tidyverse)

bike_data <- read_csv('https://raw.githubusercontent.com/dab31415/DATA606/main/Project_1/day.csv')

```


### Research question 

Is temperature an accurate predictor for bicycle ride sharing?

### Cases 

Each observation is for a single day and contains the number of bicycles rented by casual and registered users in Washington DC. The observation also contains weather statistics for temperature, humidity, wind speed, and precipitation. 

### Data collection 

Bike Sharing data was collected through https://capitalbikeshare.com which provides bicycle rentals in Washington, DC. The researched collected two years of bike sharing data then added corresponding weather and seasonal information.

### Type of study 

This is an observational study.

### Data Source 

The data was obtained at https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset, and was originally published in 2013. [@fanaee]

### Dependent Variable

The dependent variable is the number of bicycles that were rented each day.

### Independent Variable

The independent variables are temperature and weather type.

### Relevant summary statistics 


```{r}
summary(bike_data)

bike_data %>%
  ggplot(aes(temp,cnt)) + 
  geom_point(aes(color = weathersit)) + 
  geom_smooth(method = lm) +
  facet_wrap(~ weathersit)
  
```

